---
title: 'Homework: Stochastic Block Model'
author: 'MAP573 team'
date: "11/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
set.seed(1234)
```

## Packages

First, make sure you have the following packages installed:

```{r, eval=FALSE}
install.packages("igraph")
install.packages("sna")
install.packages("blockmodels")
```

Let us load the package [igraph](https://igraph.org/r/) which is particularly useful to manipulate graph data.

```{r, eval = FALSE}
library(igraph)
```

## Graph manipulation

Let us load the [Zachary's Karate Club dataset](http://aris.ss.uci.edu/~lin/76.pdf). 

It is composed of links between 34 members of a karate club. Given the club's small size, each club member knew everyone else. Sociologist Wayne Zachary documented pairwise links between members who regularly interacted outside the club.

<br>

Conflit and split:

- A conflict between the club's president and the instructor split the club into two.

- About half of the members followed the instructor ad the other half the president, a breakup that unveiled the ground truth, representing club's underlying community structure.

- Today community finding algorithms are often tested based on their ability to infer these two communities from the structure of the network before the split.

```{r, eval = FALSE}
g <- igraph::read.graph(file="data/karate.gml", format="gml")
```

<br>

1) Give a quick summary of the graph (plot, number of vertices, edges, id associated to each vertex).

<br>

2) Who are the most 'important' nodes?


<br>


3) Study the empirical degree distribution of the nodes. Use a Gaussian kernel for the non-parametric estimation of the degree distribution. Comment.

<br>

4) Remove the node with $id=1$ and illustrate the effect on the resulting network.


<br>

## Graph visualisation

We now load the package [sna](https://cran.r-project.org/web/packages/sna/sna.pdf) which is particularly useful to visualise networks.

```{r, results="hide", message=FALSE, eval = FALSE}
library(sna)
```

<br>

1) Test the different visualisation algorithms of sna (look at the _mode_ parameter). Rely on the function gplot.
Note that most functions of _sna_ work with adjacency matrices.

<br>

## Graph models: the simulation part

Let us now simulate various random graph models.

1) Simulate an undirected graph (without self-loops) with $n=50$ nodes from an ER model with $p=0.1$.

<br>

2) Simulate an undirected graph (without self-loops) with $n=50$ nodes from a SBM model with $K=3$, the same cluster proportions, and a matrix $\pi$ such that $\pi_{kk}=0.9,\forall k$ and $\pi_{kl}=0.1,\forall k \neq l$.

<br>


## Graph models: analysis of real data

We now load the package [blockmodels](https://cran.r-project.org/web/packages/blockmodels/blockmodels.pdf) to perform inference in the SBM model. 

```{r, results="hide", message=FALSE}
library(blockmodels)
```

<br>

1) Analyse the karate data set with the spectral clustering algorithms that you implemented in last homework.

<br>

2) Analyse the karate data set with blockmodels. Plot the ICL criterion w.r.t. the number of clusters ranging from $1$ to $6$.

<br>

3) Using the *cluster_edge_betweenness()* and the *dendPlot(, mode="hclust")* functions from the *igraph* package, detect communities based on edge betweenness. Plot the results.

<br>

4) Compare the obtained classification. Comment

<br>
