---
title: "MAP 573 - Data Analysis and Unsupervised Learning"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readODS)
library(DT)
```

## Course description

The objective of MAP 573 is to give students a comprehensive introduction to unsupervised analysis (downsizing and clustering) and to acquire solid and practical skills for the exploratory analysis of current data sets, using R software.

The course begins with two sessions presenting the basics of programming with the R language, as well as data manipulation and the tidyverse graphical representation libraries. R's Python interfacing capabilities are also discussed.

The course then presents the classical methods of dimensionality reduction and clustering in detail (PCA, mixture models, k-means, CAH), with a brief introduction to the many approaches that generalize them (kernel approaches, embedding, etc.).

Finally, the third part of the module proposes to the students a series of data sets to be studied in groups using the methods of the course and their generalizations. These projects will constitute the main part of the evaluation of this module.


## Schedule

  - **Teachers** : Julien Chiquet (coordinator), Florian Bourgey, Élise Dumas
  - **Lectures/tutorials are 100% online** via zoom ([link in moodle](https://moodle.polytechnique.fr/enrol/index.php?id=9404)).  Amphi Gay-Lussac is available if you need 
  - **Practicals and projects follow-up are hybrid** (face-to-face in PC + zoom) ([link in moodle](https://moodle.polytechnique.fr/enrol/index.php?id=9404))

```{r, echo = FALSE}
readODS::read_ods(path = "schedule.ods") %>% 
  datatable(options = list(pageLength = 24, dom = "t", ordering = FALSE), rownames = FALSE)
```

## Prerequisites

### Installing R

[Go to the CRAN webpage](https://cran.r-project.org/), select your OS and follow the instructions.

### Installing RStudio Desktop

[Go to the download page](https://rstudio.com/products/rstudio/download/#download). Select, download and install the file corresponding to your OS.

### Installing R packages

Launch Rstudio and execute the following commands in the console (at least these R packages will be needed during MAP573)

```{r, eval = FALSE}
install.packages("remotes") 
install.packages("learnr") 
install.packages("tidyverse") 
```

- **On Windows**

You may need Rtools ([dedicated page](https://cran.r-project.org/bin/windows/Rtools/)) and git ([dedicated page](https://git-scm.com/download/win))
  
- **On MacOS**

You may need XCode: [visit the dedicated page](https://mac.r-project.org/tools/), download the Mandatory tools and install them on your computer
  
- **On Linux**

If installation of a package fails in Rstudio, just _READ THE MESSAGES_: you may be asked to install some missing system libraries with, e.g.,

```{bash, eval = FALSE}
sudo apt-get install lib-missing
```


