---
title: "Installation instructions"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup instructions {.allowframebreaks}

**R** and **RStudio** are separate downloads and installations

- R is the underlying statistical computing environment

- RStudio is a graphical integrated development environment (IDE)

### Installing R

[Go to the CRAN webpage](https://cran.r-project.org/), select your OS and follow the instructions.

### Installing RStudio Desktop

[Go to the download page](https://rstudio.com/products/rstudio/download/#download). Select, download and install the file corresponding to your OS.

### Installing R packages

Launch Rstudio and execute the following commands in the console (at least these R packages will be needed during MAP573)

```{r, eval = FALSE}
install.packages("remotes")
install.packages("learnr")
install.packages("tidyverse")
```

- **On Windows**

You may need Rtools ([dedicated page](https://cran.r-project.org/bin/windows/Rtools/)) and git ([dedicated page](https://git-scm.com/download/win))

- **On MacOS**

You may need XCode: [visit the dedicated page](https://mac.r-project.org/tools/), download the Mandatory tools and install them on your computer

- **On Linux**

If installation of a package fails in Rstudio, just _READ THE MESSAGES_: you may be asked to install some missing system libraries with, e.g.,

```{bash, eval = FALSE}
sudo apt-get install lib-missing
```
