---
title: "Introduction to R: tutorials"
---

During the first two weeks of MAP573, students are introduced to the basics of the R language: R-base programming, data manipulation, visualization.

To do so, the lectures are accompanied by interactive tutorials which are shamelessly taken from [rstudio-education](https://rstudio.cloud/learn/primers).

These tutorials are either available online (via [a shiny server](http://shiny.cremeriefamily.info)) or as an R package so that you can install them on you computer and follow the tutorials without access to an internet connection. Prefer this last solution.

## R package MAP573tutorials

First, you need the **remotes** package to install `R` packages from other sources than the CRAN (install just once!)

```{r install remotes, eval = FALSE}
install.packages("remotes")
```

Then, you can instakl the `R` packages that contains the `R` tutorials of MAP573:

```{r install tuto package, eval = FALSE}
remotes::install_github("jchiquet/MAP573", subdir = 'MAP573tutorials')
```

Once this package is installed, you can access the tutorials as follows (only one at a time):

### R Basics

```{r launch R base tuto, eval = FALSE}
learnr::run_tutorial("basics_prog", package = "MAP573tutorials")
learnr::run_tutorial("basics_vis" , package = "MAP573tutorials")
```

### Data manipulation

```{r launch manip tuto, eval = FALSE}
learnr::run_tutorial("transform_structure", package = "MAP573tutorials")
learnr::run_tutorial("transform_extract"  , package = "MAP573tutorials")
learnr::run_tutorial("transform_summarise", package = "MAP573tutorials")
```

### Data visualization

```{r launch vis tuto, eval = FALSE}
learnr::run_tutorial("vis_01-exploratory-data-analysis", package = "MAP573tutorials")
learnr::run_tutorial("vis_02-bar-charts"               , package = "MAP573tutorials")
learnr::run_tutorial("vis_03-histograms"               , package = "MAP573tutorials")
learnr::run_tutorial("vis_04-boxplots"                 , package = "MAP573tutorials")
learnr::run_tutorial("vis_05-scatterplots"             , package = "MAP573tutorials")
learnr::run_tutorial("vis_06-line-graphs"              , package = "MAP573tutorials")
learnr::run_tutorial("vis_07-overplotting"             , package = "MAP573tutorials")
learnr::run_tutorial("vis_08-customize"                , package = "MAP573tutorials")
```

## Shiny server (limited access)

This runs on my personal server, so prefer the package solution.

  1. **R Basics**
        - [Base R fundamentals](http://cremeriefamily.info:8080/MAP573/basics_prog)
        - [Visualization fundamentals](http://cremeriefamily.info:8080/MAP573/basics_vis)
  2. **Data manipulation**
        - [The 'tibble' Structure](http://cremeriefamily.info:8080/MAP573/transform_structure)
        - [Extract from data](http://cremeriefamily.info:8080/MAP573/transform_extract)
        - [Summarise data](http://cremeriefamily.info:8080/MAP573/transform_summarise)
  3. **Vizualisation**
        - [Exploratory analysis](http://cremeriefamily.info:8080/MAP573/vis_01-exploratory-data-analysis)
        - [Bar charts](http://cremeriefamily.info:8080/MAP573/vis_02-bar-charts)
        - [Histograms](http://cremeriefamily.info:8080/MAP573/vis_03-histograms)
        - [Boxplots](http://cremeriefamily.info:8080/MAP573/vis_04-boxplots)
        - [Scatterplots](http://cremeriefamily.info:8080/MAP573/vis_05-scatterplots)
        - [Line graphs](http://cremeriefamily.info:8080/MAP573/vis_06-line-graphs)
        - [Overplotting](http://cremeriefamily.info:8080/MAP573/vis_07-overplotting)
        - [Customize](http://cremeriefamily.info:8080/MAP573/vis_08-customize)

