---
title: "Tutorial: Gaussian mixture model and the EM algorithm"
author: "MAP573 -- Introduction to unsupervised learning"
date: "École Polytechnique - Autumn 2019"
fontsize: 11pt
lang: en
geometry: left=1.45in,top=1.35in,right=1.45in,bottom=1.35in
classoption: a4paper
linkcolor: red
urlcolor: blue
citecolor: green
output:
  pdf_document:
    number_sections: true
    citation_package: natbib
    includes:
      in_header: ../preamble.sty

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
```

# Preliminaries {-}

\textsf{Goals.}

1. Gaussian mixture models
2. Expectation-Maximization algorithm for mixture models

\textsf{Instructions.} Each student _may_ send an `R markdown` report generated via `R studio` to <julien.chiquet@inra.fr> at the end of the tutorial. This report should answer the questions by commentaries and codes generating appropriate graphical outputs. [A cheat sheet of the  markdown syntax can be found here.](https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet)

\textsf{Required packages.} Check that the following packages are available on your computer:
```{r, message=FALSE}
library(aricode)
library(mixtools)
```

You also need `Rstudio`, \LaTeX\ and packages for markdown: 
```{r, message=FALSE}
library(knitr)
library(rmarkdown)
```

# Gaussian Mixture Models

We    consider   a   collection   of    random   variables $(X_1, \dots,  X_n)$ associated  with $n$  individuals drawn from $Q$ populations. The label of each individual describes the population (or class) to  which it belongs and  is unobserved.  The $Q$  classes have _a priori_ distribution ${\boldsymbol     \alpha} = (\alpha_1,\dots,\alpha_Q)$ with
$\alpha_q  = \mathbb{P}(i  \in  q)$.  The  hidden  random indicator  variables $(Z_{iq})_{i\in\mathcal{P},  q\in\mathcal{Q}}$ describe  the label  of each individuals, that is,

\begin{equation*}
  \label{eq:prior_classes}
  \alpha_q  = \mathbb{P}(Z_{iq}  =  1) =  \mathbb{P}(i  \in q),  \quad
  \text{ such that} \sum_{q=1}^Q \alpha_q = 1.
\end{equation*}
Remark that we have $\mathbf{Z}_i      =      (Z_{i1},       \dots,      Z_{iQ})      \sim
\mathcal{M}(1,\boldsymbol\alpha)$. The distribution of  $X_i$ conditional on the label of  $i$ is assumed
to be a univariate gaussian distribution with unknown parameters, that
is, $X_i | Z_{iq} = 1 \sim \mathcal{N}(\mu_q,\sigma^2_q)$.

# Questions

- _Likelihood._ Write the model complete-data loglikelihood.
- _E-step._ For fixed values of  $\hat\mu_q, \hat\sigma_q^2$ and  $\hat\alpha_q$, give  the expression  of the  estimates of the posterior probabilities $\tau_{iq}= \mathbb{P}(Z_{iq}=1|X_i)$.
- _M-step._  For fixed values of  $\hat\tau_{iq}$, show that the maximization step leads to the following estimator for the model parameters:
\begin{equation*}
  \hat{\alpha}_q =  \frac{1}{n} \sum_{i=1}^n  \hat{\tau}_{iq}, \quad
  \hat\mu_q     =    \frac{\sum_i     \hat{\tau}_{iq}    x_i}{\sum_i
  \hat{\tau}_{iq}},    \quad   \hat{\sigma}^2_q    =   \frac{\sum_i
  \hat{\tau}_{iq} (x_i-\hat\mu_q)^2}{\sum_i \hat{\tau}_{iq}}
\end{equation*}
- _Implementation._  Test  your  EM  algorithm on  simulate data.  Try  different values for $\mu_q$,  $\sigma_q$. Also consider different initialization. Compare with the output of the function \texttt{normalmixEM} in the package \textbf{mixtools}.
- _Model Selection._ Compute  the ICL criterion and test  it on your simulated data.
\begin{equation*}
  ICL(Q) =-2 \log L(X,\hat{Z}; \hat{\alpha},\hat{\mu}, \hat{\sigma^2}) + \log(n) \mathrm{df}(Q).
\end{equation*}
