---
title: "Tutorial: Gaussian mixture model and the EM algorithm"
author: "MAP573 -- Introduction to unsupervised learning"
date: "École Polytechnique - Autumn 2019"
fontsize: 11pt
lang: en
geometry: left=1.45in,top=1.35in,right=1.45in,bottom=1.35in
classoption: a4paper
linkcolor: red
urlcolor: blue
citecolor: green
output:
  pdf_document:
    number_sections: true
    citation_package: natbib
    includes:
      in_header: ../preamble.sty

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
set.seed(1234)
```

# Gaussian Mixture Models

We    consider   a   collection   of    random   variables $(X_1, \dots,  X_n)$ associated  with $n$  individuals drawn from $Q$ populations. The label of each individual describes the population (or class) to  which it belongs and  is unobserved.  The $Q$  classes have _a priori_ distribution ${\boldsymbol     \alpha} = (\alpha_1,\dots,\alpha_Q)$ with
$\alpha_q  = \mathbb{P}(Z_i  \in  q)$. In other word, the latent variable $Z_i \in \{1,..,Q\}$ indicating the label follow a multinomial distribution $Z_i \sim \mathcal{M}(1,\boldsymbol\alpha)$, such as $\sum_{q=1}^Q \alpha_q = 1$.

The distribution of  $X_i$ conditional on the label of  $i$ is assumed
to be a univariate gaussian distribution with unknown parameters, that
is, $X_i | Z_i = q \sim \mathcal{N}(\mu_q,\sigma^2_q)$.

Hence, the marginal likelihood can be written as
\begin{equation}
  \label{eq:likelihood}
  L(X) = p(X) = \prod_{i=1}^n p(X_i) = \prod_{i=1}^n \sum_{q=1}^Q  p(X_i, Z_i = q) = \prod_{i=1}^n \sum_{q=1}^Q p(Z_i = q) p(X_i | Z_i = q)
\end{equation}

Equivalently, the loglikelihood is 
\begin{equation*}
  \log L(X)  = \sum_{i=1}^n \log \sum_{q=1}^Q p(Z_i = q) p(X_i | Z_i = q)
\end{equation*}

## Alternative (equivalent) formulation {-}

Introducing the additional notation $Z_{iq} = \1_{\{Z_i = q\}}$, then the following writting will be useful:
\begin{equation*}
  \log p(Z_i) = \log\left(\prod_{q=1}^Q (\alpha_q)^{Z_{iq}}\right)= \sum_{q=1}^Q Z_{iq} \log (\alpha_q), 
\end{equation*}
which is non zero just when $q$ is the effective realization of $Z_i$. Similarly \footnote{We denote by $\mathcal{N}(X_i;\mu_\ell,\sigma_\ell)$ the density of a normal distribution at point $X_i$, with a slight notational abuse.}, 
\begin{equation*}
  \log p(X_i | Z_i) = \log\left(\prod_{q=1}^Q (\mathcal{N}(X_i;\mu_q,\sigma_q))^{Z_{iq}}\right) = \sum_{q=1}^Q Z_{iq} \log (\mathcal{N}(X_i;\mu_q,\sigma_q)).
\end{equation*}

Hence, the complete data likelihood can be written
\begin{equation*}
  L(X, Z) = p(X, Z) = \prod_{i=1}^n p(X_i, Z_i) =  \prod_{i=1}^n p(Z_i) p(X_i | Z_i) = \prod_{i=1}^n \prod_{q=1}^Q (\alpha_q \mathcal{N}(X_i;\mu_q,\sigma_\ell))^{Z_{iq}},
\end{equation*}
and then
\begin{equation}
  \label{eq:cloglikelihood}
  \log L(X, Z) =  \sum_{i=1}^n  \sum_{q=1}^Q Z_{iq} \log (\alpha_q \mathcal{N}(X_i;\mu_q,\sigma_q)),
\end{equation}

\textbf{Keep in mind that the difference between the marginal likelhood and the complete likelhood is that the $Z_i$ are given in the latter, not in the former.} Then in expression \eqref{eq:likelihood} the summatio over the class is indeed an integration over all possible values, while in \eqref{eq:cloglikelihood} all element in $\sum_q$ will be zero but the one truly observed  (the $q$ such that $Z_{iq} = 1$).

# Questions


## _Likelihood._ Write the model complete-data loglikelihood.

We denote the vector of parameters to be estimated by
$\mathbf{\mu} = (\mu_1,\dots,\mu_Q)$, $\mathbf{\sigma^2} =
(\sigma^2_1,\dots,\sigma^2_Q)$, $\boldsymbol\tau = (\tau_{iq,
i=1,\dots,n; q=1,\dots Q})$. The negative complete-data
loglikelihood is derived as follows

\Answer{%
\begin{equation*}
\begin{aligned}
\log L(\boldsymbol{\mu},\boldsymbol{\sigma}^2,\boldsymbol\tau; \mathbf{x}) & = \sum_{i=1}^n \log \left(\prod_{q=1}^Q \alpha_q f(x_i; \mu_q,\sigma^2_q)^{Z_{iq}} \right)\\
& = \sum_{i=1}^n \sum_{q=1}^Q Z_{iq} \left(\log  \frac{\alpha_q}{\sigma\sqrt{2\pi}} \exp\{\frac{1}{2\sigma^2}(x_i - \mu_q)^2\} \right)\\
& = \sum_{i=1}^n \sum_{q=1}^Q Z_{iq} \left(\log \alpha_q - \log\sigma -\log(\sqrt{2\pi}) - \frac{1}{2\sigma_q^2} (x_i - \mu_q)^2 \right)
\end{aligned}
\end{equation*}
}

## _E-step._ 

For fixed values of  $\hat\mu_q, \hat\sigma_q^2$
and  $\hat\alpha_q$, give  the expression  of the  estimates of  the
posterior probabilities $\tau_{iq}= \mathbb{P}(Z_{iq}=1|X_i)$.

\Answer{
\begin{equation}
\tau_{iq} = \frac{\hat\alpha_q f(x_i; \hat{\mu}_q, \hat\sigma_q^2)}{\sum_{q=1}^Q \hat\alpha_q f(x_i; \hat{\mu}_q, \hat\sigma_q^2)},
\end{equation}
where $f$ is the density of the normal distribution.
}

## _M-step._ 

\Answer{ The maximization step consists in solving the following optimization problem
\begin{equation}
\label{eq:m_step_optim}
\arg \max_{\sigma_q,\mu_q,\alpha_q}  \underbrace{\sum_{i=1}^n \sum_{q=1}^Q \hat\tau_{iq} \left(\log \alpha_q - \log\sigma -\log(\sqrt{2\pi}) - \frac{1}{2\sigma_q^2} (x_i - \mu_q)^2 \right)}_{Q(\boldsymbol\sigma,\boldsymbol\mu,{\boldsymbol\alpha};\hat{\boldsymbol\tau})}
\end{equation}

Consider first the mixture coefficients. We solve the above
maximization problem under the constraint that the mixture
coefficients sum to 1. This can be dealt with the Lagrange
multiplier technique. By deriving the objective function w.r.t
$\alpha_q$, we get

\begin{equation} 
\frac{\sum_i\tau_{iq}}{\alpha_q} + \lambda = 0 \Leftrightarrow  \alpha_q = \frac{\sum_i\tau_{iq}}{-\lambda}  
\end{equation}

where $\lambda$ corresponds to the Lagrange multiplier
associated with the constraint $\sum_q \alpha_q=1$. If we sum
the latter result over all $q$, we get that $1 = \sum_q
\tau_{iq} / (-\lambda)$. In other words, $\lambda = -\sum_q
\tau_{iq}$ so that finally
\begin{equation} 
\hat\alpha_q = \frac{\sum_{i=1}^n \tau_{iq}}{\sum_{i=1}^n\sum_{q=1}^Q \tau_{iq}}  
\end{equation}

Concerning, $\hat\mu_q$, null gradient condition leads to 
\begin{equation} 
\sum_i \frac{\tau_{iq}}{2\sigma_q^2}(x_i-\mu_q)  = 0 \Leftrightarrow  \mu_q = \frac{\sum_i \tau_{iq} x_i}{\sum_i \tau_{iq}}
\end{equation}

Similarly, for $\hat\sigma_q$, we get
\begin{equation} 
\sum_{i=1}^n \tau_{iq} \left(-\frac{1}{2\sigma_q^2} + \frac{1}{2\sigma_q^4}(x_i-\mu_q)^2 \right)  = 0 \Leftrightarrow  \sigma^2_q = \frac{\sum_{i=1}^n \tau_{iq} (x_i-\mu_q)^2}{\sum_{i=1}^n \tau_{iq}}
\end{equation}

}

- _Implementation._  

```{r auxiliary functions}
get_cloglik <- function(X, Z, theta) {
  n <- length(x); Q <- ncol(Z)
  alpha <- theta$alpha; mu <- theta$mu; sigma <- theta$sigma
  xs <- scale(matrix(X, length(X), length(alpha)), mu, sigma)
  log_alpha <- matrix(rep(log(alpha), n), n, Q, byrow = TRUE)
  log_sigma <- matrix(rep(log(sigma), n), n, Q, byrow = TRUE)
  res <- sum(Z*(log_alpha - log_sigma - .5*(log(2*pi) + xs^2)))
  res
}

M_step <- function(X, tau) {
  n <- length(X); Q <- ncol(tau)
  alpha  <- colMeans(tau)
  mu     <- colMeans(tau * matrix(X,n,Q)) / alpha
  sigma  <- sqrt(colMeans(tau*sweep(matrix(X,n,Q),2,mu,"-")^2)/alpha)
  list(alpha = alpha, mu = mu, sigma = sigma)
}

E_step <- function(X, theta) {
  probs <- mapply(function(alpha, mu, sigma) {
    alpha*dnorm(X,mu,sigma)
  }, theta$alpha, theta$mu, theta$sigma)
  likelihoods <- rowSums(probs)
  list(tau = probs / likelihoods, loglik = sum(log(likelihoods)))
}

```

```{r EM}
EM_mixture <- function(X, Q,
                       init.cl = sample(1:Q,n,rep=TRUE), max.iter=100, eps=1e-5) {
    n <- length(X); tau <- matrix(0,n,Q); tau[cbind(1:n,init.cl)] <- 1
    loglik  <- vector("numeric", max.iter)
    Eloglik <- vector("numeric", max.iter)
    iter <- 0; cond <- FALSE

    while (!cond) {
        iter <- iter + 1
        ## M step
        theta <- M_step(X, tau)
        ## E step
        res_Estep <- E_step(X, theta)
        tau <- res_Estep$tau
        ## check consistency
        loglik[iter]  <- res_Estep$loglik
        Eloglik[iter] <- get_cloglik(X, tau, theta)
        if (iter > 1)
            cond <- (iter>=max.iter) | Eloglik[iter]-Eloglik[iter-1] < eps
    }

    res <- list(alpha = theta$alpha,  mu = theta$mu,  sigma = theta$sigma,
                tau   = tau, cl = apply(tau, 1, which.max),
                Eloglik = Eloglik[1:iter],
                loglik  = loglik[1:iter])
    res
}
```

## Examples

We test ICL and BIC on a simple example with 4 groups

Let us start with the data generation.

```{r data generation}
mu1 <- 5   ; sigma1 <- 1; n1 <- 100
mu2 <- 10  ; sigma2 <- 1; n2 <- 200
mu3 <- 15  ; sigma3 <- 2; n3 <- 50
mu4 <- 20  ; sigma4 <- 3; n4 <- 100
cl <- rep(1:4,c(n1,n2,n3,n4))
x <- c(rnorm(n1,mu1,sigma1),rnorm(n2,mu2,sigma2),
       rnorm(n3,mu3,sigma3),rnorm(n4,mu4,sigma4))
n <- length(x)

## we randomize the class ordering
rnd <- sample(1:n)
cl <- cl[rnd]
x  <- x[rnd]

alpha <- c(n1,n2,n3,n4)/n

curve(alpha[1]*dnorm(x,mu1,sigma1) +
      alpha[2]*dnorm(x,mu2,sigma2) +
      alpha[3]*dnorm(x,mu3,sigma3) +
      alpha[4]*dnorm(x,mu4,sigma3), 
      col="blue", lty=1, from=0,to=30, n=1000,
      main="Theoretical Gaussian mixture and its components",
      xlab="x", ylab="density")
curve(alpha[1]*dnorm(x,mu1,sigma1), col="red", add=TRUE, lty=2)
curve(alpha[2]*dnorm(x,mu2,sigma2), col="red", add=TRUE, lty=2)
curve(alpha[3]*dnorm(x,mu3,sigma3), col="red", add=TRUE, lty=2)
curve(alpha[4]*dnorm(x,mu4,sigma4), col="red", add=TRUE, lty=2)
rug(x)
```

Suppose that we know the number of components, i.e. 4.

```{r test EM}
out <- EM_mixture(x, Q = 4)
plot(out$loglik, main = "data log-likelihood", type="l")

curve(alpha[1]*dnorm(x,mu1,sigma1) +
      alpha[2]*dnorm(x,mu2,sigma2) +
      alpha[3]*dnorm(x,mu3,sigma3) +
      alpha[4]*dnorm(x,mu4,sigma3), col="blue",
      lty=1, from=0,to=30, n=1000,
      main="Theoretical Gaussian mixture and estimated components",
      xlab="x", ylab="density")
curve(out$alpha[1]*dnorm(x,out$mu[1],out$sigma[1]), col="red", add=TRUE, lty=2)
curve(out$alpha[2]*dnorm(x,out$mu[2],out$sigma[2]), col="red", add=TRUE, lty=2)
curve(out$alpha[3]*dnorm(x,out$mu[3],out$sigma[3]), col="red", add=TRUE, lty=2)
curve(out$alpha[4]*dnorm(x,out$mu[4],out$sigma[4]), col="red", add=TRUE, lty=2)
rug(x)

## the confusion table gives rather good results
table(out$cl,cl)
```

The number of component mixture is hard to recover because of the last two mixed components

```{r model selection}
seq.Q <- 2:6
crit.EM <- sapply(seq.Q, function(Q) {
    out <- EM_mixture(x, Q, kmeans(x,Q)$cl)
    df <- Q - 1 + 2 * Q
    return(c(BIC = -2*tail(out$loglik ,1) + log(n)*df,
             ICL = -2*tail(out$Eloglik,1) + log(n)*df ))
})
matplot(seq.Q, t(crit.EM), type="l", col=c("red", "blue"))
legend("topleft", c("BIC", "ICL"), col=c("red", "blue"),lty=1)

Q.hat <- seq.Q[which.min(crit.EM[1, ])]
out <- EM_mixture(x, Q = Q.hat, kmeans(x, Q.hat)$cl)
par(mfrow=c(1,1))
curve(alpha[1]*dnorm(x,mu1,sigma1) +
      alpha[2]*dnorm(x,mu2,sigma2) +
      alpha[3]*dnorm(x,mu3,sigma3) +
      alpha[4]*dnorm(x,mu4,sigma3), col="blue", 
      lty=1, from=0,to=30, n=1000,
      main="Theoretical Gaussian mixture and estimated components",
      xlab="x", ylab="density")
for (q in 1:Q.hat) {
    curve(out$alpha[q]*dnorm(x,out$mu[q],out$sigma[q]), col="red", add=TRUE, lty=2)
}
rug(x)
```
